---
title: "Campaign Finance"
output: 
html_notebook: default
---

### Introduction

Each candidate for federal office (U.S. House of Representatives, U.S. Senate, President) must register with the Federal Election Commission (FEC) and form a Political Action Committee (PAC) to receive contributions. Candidates can have several different kinds of PACs, but everyone has a "principal campaign committee" through which they receive money.

It's an election year: every Representative is up for re-election in an even year, since Reps serve 2-year terms. One third of the Senate is up for re-election every two years because Sens serve 6-year terms. Candidates can receive money even if they're not running for re-election, but though there are limits to how much any entity can give directly to a candidate. There are no limits for receiving or spending money as independent expenditures, which support or oppose a candidate but don't coordinate with that candidate.

Missouri has 8 House seats and 2 Senate seats; only 1 Senate seat is up for re-election in 2022; Roy Blunt is not running for re-election so it's an open seat. Josh Hawley won't be up for re-election until 2024; however senators are always raising money whether they're up for re-election in the current cycle or not.

### Documentation

-   [Committee donations & independent expenditures](https://www.fec.gov/campaign-finance-data/contributions-committees-candidates-file-description/)
-   [Candidates](https://www.fec.gov/campaign-finance-data/candidate-master-file-description/)
-   [Committees](https://www.fec.gov/campaign-finance-data/committee-master-file-description/)

### Setup

Load your packages

```{r}
library(tidyverse)
library(lubridate)
```

Load the data tables

```{r}

# A note: I want to coerce some of the columns in the "transactions" table into character columns, so I'm using the col_types argument within read_csv. You can list data types for each column. Notice the one d, which stands for "double", or number, in the 15th position. This is for the transaction_amt column.

transactions <- read_csv("https://github.com/eklucas/ADJF2022/raw/main/data/campfin/transactions.csv", col_types=c("ccccccccccccccdccccccc"))

transaction_types <- read_csv("https://github.com/eklucas/ADJF2022/raw/main/data/campfin/transaction_types.csv")
committees <- read_csv("https://github.com/eklucas/ADJF2022/raw/main/data/campfin/committees.csv")
candidates <- read_csv("https://github.com/eklucas/ADJF2022/raw/main/data/campfin/candidates.csv")
```

Campaign finance data, like many datasets, comes in relational tables. The FEC stores information about candidates and committees in separate files from the main contributions data to save space. In order to fully understand the data you need to join these tables together to access the important information in each. There's also a lookup table in here that describes the transaction types, which are important to pay attention to. For example: the transaction type "24A" indicates that the amount is an independent expenditure in opposition of the noted candidate. "24E" is an expenditure in support of the noted candidate.

### Get to know the data

I'll offer up some pointers on how this data works, but you should do some exploring on your own as well.

Start by looking at the different transaction types in this dataset, which is a slice of the main data.

```{r}
transactions %>% count(transaction_tp)
```

There are only a handful of transaction types in here: what are they? To find out, we could look them up in the `transaction_types` table, or we could bring in the `transaction_desc` field from that table, using a join.

Which type of join? We don't care about all the transaction types in t the transaction types that are in our main table. So we'll use a **left_join** (returns every row from table x and only matches from table y). That table, onlyhe shared column in these two tables is `transaction_tp`.

```{r}
transactions %>% 
  count(transaction_tp) %>%
  left_join(transaction_types, by="transaction_tp")
```

Most of these transactions are "Contributions made to nonaffiliated committees", which basically means PACs that are run by companies, labor unions, and special interest groups giving money to candidate committees. There are some independent expenditures in here as well.

Next look at which candidates are included in this dataset. I said it was a slice of the main data; I only included money going to certain candidates. The `transactions` table has a cand_id, an id that is unique for each candidate running for office. But we can't tell who the person is by looking at the id, so we need to join to the `candidates` table.

Which type of join? Let's start with the candidates table and see which candidates match the cand_ids in our `transactions` table. We don't yet care about the information in transactions, so we'll use a semi_join:

```{r}
candidates %>% semi_join(transactions, by="cand_id")
```

35 candidates are in our data. What characterizes them?

```{r}
# Which states are they running in?
candidates %>% 
  semi_join(transactions, by="cand_id") %>% 
  count(office_st)

# How many are vying for the House vs the Senate?
candidates %>% 
  semi_join(transactions, by="cand_id") %>% 
  count(office)

# How many in each party?
candidates %>% 
  semi_join(transactions, by="cand_id") %>% 
  count(party)

```

Finally, who is spending the money? These transaction types all indicate spending rather than receiving money. So the committees spending the money are in the `cmte_id` field of the `transactions` table. But to find out who they are, we need the `committees` table. Write a join to find out which committees are in the data:

```{r}
committees %>% 
  semi_join(transactions, by="cmte_id") %>%
  count(cmte_nm)
```

One final note about the data: the `name` column in the transactions table points to different entities depending on the transaction type. For independent expenditures (24A and 24E), the `name` is the vendor who received the money for the ad or whatever the committee spent money on to support or oppose a candidate. For the other committee contributions (24K) the `name` is the committee that received the money. For in-kind contributions (24Z) it can be the candidate receiving the item (things that have monetary value but are not actual money), or it might be another entity receiving the item.

Now: explore the data! Try answering the questions below. You're also free to explore other things that interest you. Spend some time with the data and turn in your script by Monday.

-   Which Senate candidate is receiving the most money from committees? Which House candidate?
```{r}
candidates %>% 
  filter(office=="S") %>% 
  left_join(transactions, by="cand_id") %>% 
  group_by(cand_name) %>% 
  summarise(sum = sum(transaction_amt)) %>% 
  arrange(desc(sum))

candidates %>% 
  filter(office=="H") %>% 
  left_join(transactions, by="cand_id") %>% 
  group_by(cand_name) %>% 
  summarise(sum = sum(transaction_amt)) %>% 
  arrange(desc(sum))
```

-   Which party has the most independent expenditures spent against it? For it?
24A	- Independent expenditure opposing election of candidate
24E	- Independent expenditure advocating election of candidate
```{r}
candidates %>%
  left_join(transactions, by="cand_id") %>% 
  filter(transaction_tp == "24A") %>% 
  group_by(party) %>% 
  summarise(sum = sum(transaction_amt))

candidates %>%
  left_join(transactions, by="cand_id") %>% 
  filter(transaction_tp == "24E") %>% 
  group_by(party) %>% 
  summarise(sum = sum(transaction_amt))

candidates %>%
  right_join(transactions, by="cand_id") %>%
  filter(party == c("DEM", "REP", "IND")) %>% 
  group_by(party, transaction_tp) %>%
  summarise(sum = sum(transaction_amt)) %>% 
  pivot_wider(
    names_from = transaction_tp, 
    values_from = sum)
```

-   Create a table with one row for each candidate that lists total independent expenditures spent for and against that candidate. Is there anything interesting or surprising here?

```{r}
candidates %>%
  right_join(transactions, by="cand_id") %>% 
  group_by(cand_name, transaction_tp) %>% 
  summarise(sum = sum(transaction_amt)) %>% 
  pivot_wider(
    names_from = transaction_tp,
    values_from = sum
  )
```

-   Are there any PACs giving money to candidates that are based in Missouri? Often local companies spend money on federal elections.

```{r}
committees %>% 
  filter(cmte_st == "MO")

# I'm not sure what this asking that requires a join, so I going to answer who each committee is giving to and how much
committees %>% 
  right_join(transactions, by="cmte_id") %>% 
  filter(cmte_st == "MO") %>% 
  left_join(candidates, by=c("cand_id.y" = "cand_id")) %>% 
  group_by(cmte_nm, cand_name) %>% 
  summarise(sum = sum(transaction_amt))
  
```

