---
title: "Dispatch per capita"
output: html_notebook
---

Data is year-to-date

Population of Columbia, MO is 125,691.

```{r}
library(tidyverse)
library(lubridate)

# Read data
dispatch <- read_csv("data/dispatch.csv") %>% mutate(CallDateTime = mdy_hms(CallDateTime))
```

Crime Categories per 100,000

```{r}
# Columbia Pop.
columbia <- 125691

# Calculating per capita
dispatch_per_capita <- dispatch %>% count(ExtNatureDisplayName) %>% arrange(desc(n)) %>% mutate(DispatchCallsPerCapita = round((n/columbia) * 100000, 2)) %>% rename(DispatchCalls = n)
```

I need a full year of data to make this data comparable. So, I loaded all the data in the last year.

```{r}
# Read data
dispatch_year <- read_csv("data/dispatch_year.csv") %>% mutate(CallDateTime = mdy_hms(CallDateTime))

# Check date range
dispatch_year %>% summarise(range(CallDateTime))
```

```{r}
# Calculate per year
dispatch_year_per_capita <- dispatch_year %>% count(ExtNatureDisplayName) %>% arrange(desc(n)) %>% mutate(DispatchCallsPerCapita = round((n/columbia) * 100000, 2)) %>% rename(DispatchCalls = n)
```

#### Compare to Springfield, MO

Springfield only had data by year and no data for the current year. Columbia only had data for the last 365 days. So, this data is not apples to apples.

```{r}
# Read data
dispatch_spring <- read_csv("data/dispatch_springfield.csv") %>% select(1,10) %>% rename(DispatchCalls = lastyear)

#Clean data
dispatch_spring$DispatchCalls <- str_remove_all(dispatch_spring$DispatchCalls, "[ , ]")

#Change data type to numeric
dispatch_spring$DispatchCalls <- as.numeric(dispatch_spring$DispatchCalls)
```

Springfield pop. = 167,601

```{r}
# Springfield Pop.
springfield <- 167601

# Calculate per capita
dispatch_spring_per_capita <- dispatch_spring %>%  arrange(desc(DispatchCalls)) %>%  mutate(DispatchCallsPerCapita = round(DispatchCalls/springfield * 100000, 2))
```

Compare Columbia and Springfield

```{r}
#Uppercase all springfield event types
dispatch_spring_per_capita$`Event Type` <- toupper(dispatch_spring_per_capita$`Event Type`)

#Clean data
dispatch_spring_per_capita$`Event Type` <- str_remove_all(dispatch_spring_per_capita$`Event Type`, "[ ,]")

# Add city data
dispatch_year_per_capita$City <- "Columbia"
dispatch_spring_per_capita$City <- "Springfield"

# Standardize column names
dispatch_year_per_capita <- dispatch_year_per_capita %>% rename(Name = ExtNatureDisplayName)
dispatch_spring_per_capita <- dispatch_spring_per_capita %>% rename(Name = "Event Type")

# Remove event types without matches
dispatch_matches_columbia <- semi_join(dispatch_year_per_capita, dispatch_spring_per_capita, by=c("Name" = "Name"))
dispatch_matches_spring <- semi_join(dispatch_spring_per_capita, dispatch_year_per_capita, by=c("Name" = "Name"))

# Append tables
dispatch_match <- dispatch_matches_columbia %>% bind_rows(dispatch_matches_spring) %>% arrange(desc(DispatchCallsPerCapita))
```

Create comparative table

```{r echo=TRUE}
# Join tables
dispatch_compare <- full_join(dispatch_year_per_capita,dispatch_spring_per_capita,by= c("Name" = "Name")) %>% select(Name, DispatchCallsPerCapita.x, DispatchCallsPerCapita.y) %>% rename(Columbia = "DispatchCallsPerCapita.x") %>% rename(Springfield = "DispatchCallsPerCapita.y")

dispatch_compare
```

Make comparative graph

```{r}
bargraph <- ggplot(data = dispatch_match) +
  geom_bar(aes(x = Name,
               y = DispatchCallsPerCapita,
               fill = City),
           stat = "identity",
           position = position_dodge()) +
  theme(legend.title=element_blank()) + coord_flip()
bargraph
```
