---
title: "Homework 1"
output: html_notebook
---

I chose to answer my question using Python and R, so, I can keep up my python skills as I learn R. Feel free to ignore my Python sections. It did help with accuracy because when my code in each language disagree, I knew I did something wrong.

*LIZ: sounds like a good strategy to test your R code*

Initialize libraries

```{r}
library(tidyverse)
library(reticulate)
```

```{python}
import pandas as pd
```

Load data

```{r}
hot <- read_csv("data/hot-100-current.csv")
```

```{python}
with open("data/hot-100-current.csv", "r") as file:
  hot = pd.read_csv(file)
```

Display data

```{r}
view(hot)
view(py$hot)
```

Basic data information

-   334,400 rows, 7 columns

```{r}
hot %>% count(chart_week)
```

-   3,344 weeks of charts

-   Some band names are spelled multiple ways (e.g. "Matchbox Twenty" and "Matchbox 20"). I didn't correct for this.

Question for data

1.  Which artist had the most weeks on the chart for any of their songs?

    ```{r}
    artists <- hot %>% count(performer) %>% arrange(desc(n))
    head(artists, 3)
    ```

```{python}
artists = hot.groupby("performer")["title"].count().reset_index(name="count").sort_values("count", ascending=False)
print(artists.head(3))
```

```{r}
view(py$artists)
```

Answer: Taylor Swift had the most songs charted with 1,079 weeks of a song charting. When an artist had two songs on the charts, I counted that as two weeks.

2.  What songs spent the most weeks in the top ten during this period?

    ```{r}
    topten <- hot %>% filter(current_week <= 10) %>% count(performer, title) %>% arrange(desc(n))
    head(topten, 3)
    ```

```{python}
topten = hot[hot["current_week"] <= 10].groupby(["title", "performer"])["current_week"].count().reset_index(name = "count").sort_values("count", ascending=False)
topten.head(3)
```

Answer: Blinding Lights by Weeknd spent the most time in the top ten with 57 weeks in the top ten.

3.  What artists have had the most songs in the top 100?

```{r}
charted <- hot %>% count(performer, title) %>% count(performer) %>% arrange(desc(n))
head(charted, 10)
```

```{python}
charted = hot.groupby(["title", "performer"])["chart_week"].count().count("performer").reset_index(name = "count").sort_values("count", ascending = False)
charted.head(10)
```

Answer: Surprisingly, Glee Cast had the most songs in the top 100, with 183 chart-making songs.

4.  How many artists have charted more than five times?

*LIZ: technically your code asks more than six times*
```{r}
twoplus <- charted %>% filter(n > 6) %>% count() %>% as.numeric()
twoplus
```

```{python}
twoplus = len(charted[charted["count"] > 6].groupby("performer"))
twoplus
```

Answer: 982 artists have charted five or more times.

5.  How many charted songs mention "Christmas" in title?

```{r}
christmas <- hot %>% filter(grepl("Christmas", title)) %>% count() %>% as.numeric()
christmas
```

```{python}
christmas = len(hot[hot["title"].str.contains("Christmas")])
christmas
```

Answer: 398 songs contain Christmas in its title.

*LIZ: Really good use of code here, and I'm impressed by the side by side R and Python code. Great that you note some of the artists' names are represented inconsistently. For the last question, I appreciate the answerability of the question with regard to code, but realistically wouldn't it be a better question to ask how many charted songs are Christmas songs? Which is, of course, much harder to answer. But something to keep in mind when you're thinking like a journalist; don't just ask questions that are easy to answer.*

```{r}

```

